## Topic 5: Deeper dive into how to provide and pre-process data in `teal.modules.clinical` (15 minutes)

## How to provide data to a `teal` application? - code samples

The `data` argument in `teal::init()` is where you specify all the datasets your application will use.
There are several ways to provide data to `teal`:

- using `teal_data()`
- using `cdisc_data()` for clinical data

## Option 1: Using `teal_data()`

```{r include=FALSE}
library(teal.code) # prevent messsage in next code chunk
```

```{r, messages=FALSE}
library(teal.data)

# Create a basic teal_data object
my_data <- teal_data(
  IRIS = iris,
  MTCARS = mtcars
)

# View the structure
print(my_data)
```

## Option 2: Adding code for reproducibility

One of the key features of `teal.data` is its ability to track the code used to generate data:

```{r, messages=FALSE, warning=FALSE}
# Create teal_data with explicit code tracking
my_data <- teal_data()

# Add datasets with code
my_data <- within(my_data, {
  # Load and prepare iris data
  IRIS <- iris
  IRIS$Species <- as.factor(IRIS$Species)

  # Load and prepare mtcars data
  MTCARS <- mtcars
  MTCARS$cyl <- as.factor(MTCARS$cyl)
  MTCARS$gear <- as.factor(MTCARS$gear)
})

# View the tracked code
get_code(my_data) |>
  cat()
```

## Option 3: Using `cdisc_data()` for clinical data

For clinical trial data following CDISC standards, `teal.data` provides the specialized `cdisc_data()`
function that automatically handles common clinical data relationships.

```{r}
library(teal.data)
library(pharmaverseadam)

adsl <- pharmaverseadam::adsl
adae <- pharmaverseadae::adae

# Create cdisc_data object
clinical_data <- cdisc_data(
  ADSL = adsl,
  ADAE = adae,
  code = c(
    "adsl <- pharmaverseadam::adsl",
    "adae <- pharmaverseadae::adae"
  )
)

print(clinical_data)
```

## Introduction to `teal.data` in the context of the CDISC data standard

The `teal.data` package serves as the foundation for all data operations in `teal`. It provides a structured way to:
- Store multiple datasets in a single object
- Track the code used to create or modify data
- Define relationships between datasets

When you create a `teal_data` object, `teal` does several things behind the scenes:
1. **Data storage** - Your datasets are stored within the object
2. **Code tracking** - The creation process is recorded for reproducibility
3. **Metadata creation** - Information about datasets and their relationships is stored

## Creating `cdisc_data` objects with explicit code

```{.r}
# More explicit approach with within()
library(teal.data)
library(teal)
library(pharmaverseadam)

clinical_data <- cdisc_data()
clinical_data <- within(clinical_data, {
  adsl <- pharmaverseadam::adsl
  adae <- pharmaverseadam::adae
})
join_keys(clinical_data) <- teal.data::default_cdisc_join_keys[c("ADSL", "ADAE")]

# View the structure and join keys
print(clinical_data)
print(join_keys(clinical_data))
```

## Introduction to the concept of keys in `teal.data`

`join_keys` define how datasets relate to each other, which is crucial for:
- Proper filtering across related datasets (filtering on the parent dataset filters entries in the child dataset as well)
- Correct data merging in modules (`join_key` are used by the modules to merge datasets together)

## Automatic join key detection with `cdisc_data()`

One of the key advantages of `cdisc_data()` is automatic join key detection:

```{r}
library(pharmaverseadam)
library(teal)
library(teal.data)

adsl <- pharmaverseadam::adsl
clinical_data_1 <- teal.data::cdisc_data(ADSL = adsl, code = "ADSL <- pharmaverseadam::adsl")
print(join_keys(clinical_data_1))

clinical_data_2 <- teal.data::cdisc_data()
clinical_data_2 <- within(clinical_data_2, {
  ADSL <- pharmaverseadam::adsl
})
print(join_keys(clinical_data_2))
```

Notice the difference in the two above prints. When we pass a dataset directly to
`cdisc_data` constructor, the constructor automatically adds the `join_keys` object.
Whereas in the second example, we used `within` and our object did not automatically
receive correct `join_keys`.

## Understanding join key output

```{r}
library(pharmaverseadam)
library(teal.data)
library(teal)
# Create example with multiple relationships
complex_data <- cdisc_data(
  ADSL = pharmaverseadam::adsl,
  ADAE = pharmaverseadam::adae,
  ADEX = pharmaverseadam::adex
)
join_keys(complex_data) <- teal.data::default_cdisc_join_keys[c("ADSL", "ADAE", "ADEX")]

# Understanding the output:
# - Each line shows a relationship between two datasets
# - The arrow (->) indicates the direction (parent -> child)
# - Variables in brackets show the joining columns
# - ADSL is typically the parent (subject-level data)
# - the implicit join keys are created automatically when
#   teal.data detects a possible join through a "middleman" dataset

# Print and interpret join keys
cat("Join Keys Structure:\n")
print(join_keys(complex_data))
```

## Creating a custom `teal.data` object with custom user-defined keys - code samples

Sometimes you need to create or modify join keys manually, especially when working with non-standard data structures or when you
need custom relationships.

## Manual join keys creation

```{r}
# Create a teal_data object without automatic join detection
my_data <- teal_data(
  PATIENTS = data.frame(
    patient_id = 1:100,
    age = sample(18:80, 100, replace = TRUE),
    treatment = sample(c("A", "B"), 100, replace = TRUE)
  ),
  VISITS = data.frame(
    patient_id = rep(1:100, each = 4),
    visit_num = rep(1:4, 100),
    visit_date = seq.Date(as.Date("2024-01-01"), by = "week", length.out = 400),
    measurement = rnorm(400, 100, 15)
  ),
  EVENTS = data.frame(
    patient_id = sample(1:100, 200, replace = TRUE),
    event_type = sample(c("AE", "CM", "EX"), 200, replace = TRUE),
    event_date = sample(seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day"), 200)
  )
)

# Manually define join keys
join_keys(my_data) <- join_keys(
  join_key("PATIENTS", "VISITS", "patient_id"),
  join_key("PATIENTS", "EVENTS", "patient_id")
)

# View the join keys
print("Manual join keys:")
print(join_keys(my_data))
```

## Overwriting automatic join keys in `cdisc_data`

Sometimes the automatic join key detection in `cdisc_data()` doesn't match your specific needs. Here's how to override the default behavior:

```{r}
library(pharmaverseadam)
library(teal)
library(teal.data)

custom_keys <- join_keys(
  # Primary keys for each dataset
  join_key("ADSL", "ADSL", c("STUDYID", "USUBJID")),
  join_key("ADAE", "ADAE", c("STUDYID", "USUBJID", "AESEQ")),
  join_key("ADCM", "ADCM", c("STUDYID", "USUBJID", "CMSEQ")),

  # Relationships between datasets
  join_key("ADSL", "ADAE", c("STUDYID", "USUBJID")),
  join_key("ADSL", "ADCM", c("STUDYID", "USUBJID")),

  # Custom: Allow direct relationship between ADAE and ADCM
  # This might be useful for analyzing AEs and concomitant medications together
  join_key("ADAE", "ADCM", c("STUDYID", "USUBJID"))
)

custom_data <- cdisc_data(
  ADSL = pharmaverseadam::adsl,
  ADAE = pharmaverseadam::adae,
  ADCM = pharmaverseadam::adcm
)

# Override the automatic join keys
join_keys(custom_data) <- custom_keys

# View the custom join keys
cat("\nCustom join keys:\n")
print(join_keys(custom_data))
```
